Posterior summarisation in Bayesian phylogenetics using Tracer 1.7

### 2017-12-15 manuscript/tracer17.tex
rewrote ms. switched to a single example data set. remade figures

\author[Rambaut \textit{et~al}]{ Andrew Rambaut\,$^{1}$, Alexei J.~Drummond\,$^{2,3}$, Dong Xie\,$^{2,3}$, Marc A.~Suchard\,$^{4,5}$}

\address{
$^{1}$Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK\\
$^{2}$Department of Computer Science, University of Auckland, Auckland, NZ\\
$^{3}$Centre for Computational Evolution, University of Auckland, Auckland, NZ\\
$^{4}$Departments of Biomathematics and Human Genetics, David Geffen School of Medicine, UCLA, and \\
$^{5}$Department of Biostatistics, UCLA Fielding School of Public Health, University of California, Los Angeles, USA \\
}

@@

\section{Motivation:}
%Computational evolutionary biology, statistical phylogenetics and coalescent-based population genetics are becoming increasingly central to the analysis and understanding of molecular sequence data.
Bayesian inference of phylogeny using Markov chain Monte Carlo (MCMC) plays a central role in understanding evolutionary history and processes from molecular sequence data.
Visualising and analysing the samples from the posterior distribution is a key step in any non-trivial Bayesian inference, and it is especially important for methods that employ MCMC.
\section{Results:}
Here we present a software package \emph{Tracer} (version 1.7) for visualising and analysing the trace files generated by Bayesian MCMC runs.
The \emph{Tracer} software also provides kernel density estimation, cheap marginal likelihood estimation,
multivariate visualisation, demographic trajectory reconstruction, conditional posterior distribution summary and more.
\section{Availability:}
Tracer is open-source under the GNU lesser general public license and available at
\url{http://beast.bio.ed.ac.uk/Tracer} and \url{https://github.com/beast-dev/tracer}.

@@

Bayesian inference of phylogeny using Markov chain Monte Carlo (MCMC) \citep{rannala1996probability,mau1999bayesian} flourishes as a popular approach to uncover the evolutionary relationships among taxa, whether genes, genomes, individuals or species.
MCMC approaches generate samples of model parameter values including the phylogenetic tree drawn from their posterior distribution given molecular sequence data.
Visualising, tabulating and marginalising these samples is critical for approximating the posterior quantities of interest that one reports in a Bayesian phylogenetic analysis.
%However, the full posterior distribution is high-dimensional and often difficult to visualise.
To facilitate this task, we develop the Tracer (version 1.7) software package that parses MCMC log files of samples to interactively explore the high-dimensional posterior distribution.
%
%containing the list of parameters whose values sampled from the chain.
Tracer works automatically with sample output from
%It can be used to analyse runs of
BEAST 1 \citep{drummond2012bayesian}, BEAST 2 \citep{bouckaert2014beast2}, MrBayes \citep{ronquist2012mrbayes}, RevBayes \citep{hohna2016revbayes}, LAMARC \citep{kuhner2006lamarc}, Migrate \citep{beerli2006comparison} and possibly other MCMC programs from other domains.

(move to somewhere else) In addition, Tracer's demographic reconstruction functions are utilised to redraw the distribution of effective population sizes over time if an available demographic model was involved in the analysis.

%\begin{itemize}
%\item Importance of posterior summaries (as scientifically relevant marginalization of the full posterior distribution)
%\item Full posterior distribution is high-dimensional and (often) difficult to visualize
%\item Posterior simulator agnostic -- highly used with \textsc{MrBayes} as well
%\end{itemize}

@@

% MAS: We need to limit the number of figures
%\begin{figure}[ht]
%\centerline{
%\includegraphics[width=0.45\textwidth]{./figures/tracer.pdf}
%}
%\caption{The system design of Tracer 1.7}
%\label{fig:tracer}
%\end{figure}

Tracer is able to analyse either a single MCMC log or combine the shared traces from multiple logs as designed in Figure~\ref{fig:tracer}.
The list of traces in each log are created from the sampling parameters logged during a MCMC process, which include their statistical summaries from their values at all states logged.
The effective sample size (ESS) is an important measurement designed for a sampled parameter having continuous values, which tells the number of effectively independent draws from the posterior distribution that the Markov chain is equivalent to.

Tracer handles three trace types, also known as variable types \citep{mendenhall2012introduction}: real, integer, and categorical.
The real trace is used to handle quantitative continuous variables, which is the most common type in logged parameters.
The integer is dealing with quantitative discrete variables, and the categorical trace is for qualitative categorical variables.
% check above definition
%These three types can be automatically assigned when the log file is imported. But they can also be changed later.

The multiple traces can be selected as shown in Figure~\ref{fig:multitrace}, which will overlay the plots for the different traces allowing comparisons to be made. The multiple trace files can be similarly selected as well to compare different runs. If multiple trace files have the same trace names,  then a ``Combined'' trace will automatically appear, which can be selected as well as the individual trace files.

There are four analysis tabs to choose from either a selected trace (sampling parameter) or multiple traces:

%TODO simplify, some duplication to 1st example
\begin{itemize}
\item Estimates - this shows the mean, standard deviation, confidence intervals, ESS, and other statistics about the selected trace(s). A histogram presenting the frequency distribution over values will also be plotted for a single selected trace, such as Figure~\ref{fig:4tabs}.a. Interval bars for quantitative traces (real or integer) or violin plots for categorical traces will be drawn for a comparison, such as Figure~\ref{fig:multitrace}, if multiple traces are selected.

\item Density - this shows the Bayesian posterior density plot for the selected trace(s), if they are continuous, the kernel density estimates will be available for continuous traces, for example, in Figure~\ref{fig:4tabs}.b.

\item Joint-Marginal - this only appears if exactly two traces are chosen. It then plots one against the other to look at their joint-marginal distribution, for example, in Figure~\ref{fig:4tabs}.c.

% trace is duplicate to trace, use trajectory?
\item Trace - this shows the trajectory of the selected trace(s) against state or generation number as it can be seen in Figure~\ref{fig:4tabs}.d. It can be used to check mixing, choose a suitable burn-in and look for trends that might suggest problems with convergence.

@@

In addition to illustrating joint-marginal distribution between two continuous traces, we newly invented a TangHuLu chart to visualise the joint probability between two integer or categorical traces using coloured bubbles, as shown in Figure~\ref{fig:tanghulu}.a. The size of circle is proportional to the joint probability, the blue coloured circle is in the credible set of given a probability threshold default to 0.95, the red is in the set not in the credible set. The tile background is coloured if there is any circle, in case the very small circle wouldn't be missed out. The coloured background can also help to show the area of the credible set and non-credible set.

Furthermore, box-and-whisker plots are used to display the joint-marginal distribution between one continuous trace and one integer or categorical trace as shown in Figure~\ref{fig:tanghulu}.b.

%
\begin{figure}[ht]
%\includegraphics[width=.5\textwidth]{./figures/multitrace.png}
\subfigure[Multiple traces]{\includegraphics[width=0.46\textwidth]{./figures/multitrace.png}}
\subfigure[TangHuLu chart]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/tanghulu.pdf}}
\subfigure[boxplot]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/boxplot.pdf}}
\caption{A comparison of multiple traces}
\label{fig:multitrace}
\end{figure}
%

%\begin{figure}[ht]
%\subfigure[TangHuLu chart]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/tanghulu.pdf}}
%\subfigure[boxplot]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/boxplot.pdf}}
%\caption{Visualisation of joint-marginal distribution (a) between two integer or categorical traces using TangHuLu chart, and (b) between one continuous trace and one integer or categorical trace using box-and-whisker plot.}
%\label{fig:tanghulu}
%\end{figure}

A key feature of Tracer lies in its ability to link together samples from quantitative random variable log files and tree log files.
This linkage is not available in other popular post-processing MCMC software, such as coda \citep{plummer2006coda} and AWTY \citep{nylander2007awty}, and enables users to approximate the posterior distribution of scientifically relevant quantities that are functions of both the tree and other quantitative measures.
Demographic reconstruction through time is a prime example.

Tracer also provides demographic reconstruction resulting in a graphical plot, often applyed to reconstruct epidemic dynamics, given a MCMC log from a Bayesian framework or additionally with the tree log depending on the model.
The currently available models are constant size \citep{drummond2002estimating}, exponential growth \citep{drummond2002estimating}, logistic growth (?), Bayesian coalescent skyline \citep{drummond2005bayesian}, GMRF Bayesian skyride \citep{minin2008smooth}, and Bayesian coalescent skygrid \citep{gill2012improving}.
%For example, ``Demographic Reconstruction" performs reconstructions of the distribution for a number of models, such as constant size \citep{drummond2002estimating}, exponential growth \citep{drummond2002estimating}, and logistic growth (?). %citation?
%Besides ``Bayesian Skyline Reconstruction'' reconstructs a generalised Bayesian coalescent skyline plot (e.g. Figure~\ref{fig:flu}.a) describing demographic history under canonical scenarios \citep{drummond2005bayesian},
%``GMRF Skyride Reconstruction'' and ``SkyGrid Reconstruction'' are respectively available for a temporal smoothing method using the Gaussian Markov random field (GMRF) model  \citep{minin2008smooth} and a generalisation of the GMRF model \citep{gill2012improving}.
Besides these, ``Lineages Through Time'' (?) is also available to plot the quantiles of the number of lineages against time, which describes the change of diversification of all organisms over time. %citation?
%All these models are available in BEAST \citep{drummond2007beast}.

%TODO is it OK ?
Tracer recently offers a fairly general solution of looking at conditional posterior distributions. This, specifically, supports for Bayesian stochastic search variable selection (BSSVS) forms of model averaging, in which some parameters are only in the likelihood when their sub-model is indicated by an indicator function usually assigned to a discrete, integer or boolean variable. In that case, the posterior of the parameter should not include the states when it was sampled only in the prior, because the sub-model it belonged to was ``turned off". This is relevant for extended Bayesian skyline plot, random local clock model, microsatellite model averaging, and relaxed clock model averaging methods, etc. It is also relevant for BSSVS in phylogeography as well depending on how the state is logged.

% TODO more ?

@@

%Even users of \textsc{MrBayes} and \textsc{RevBayes} find Tracer handy.

\subsection*{Calibrating the tree to estimate TMRCA}

Figure~\ref{fig:4tabs} demonstrates a snapshot of posterior summaries in Tracer for a BEAST \citep{drummond2012bayesian} run created from a 898 bp alignment of sequences of 12 species of primates (http://beast.bio.ed.ac.uk/tutorials?). %data citation

%using HKY + $\Gamma_4$ model, uncorrelated lognormal relaxed molecular clock, and Yule prior on the tree.
In this analysis, we furthermore specified three calibrations on the phylogenetic tree to estimate time to the most recent common ancestor (TMRCA) based on our prior fossil knowledge.

The first TMRCA prior, called ``ingroup'', includes all species except the \textit{Lemur} which will form the outgroup in the tree. %monophyletic
The second group is ``Human-Chimp'' that contains only \textit{Homo sapiens} and \textit{Pan}.
The last consists of everything except \textit{Lemur}, \textit{Saimiri} and \textit{Tarsius} to create the hominoid-cercopithecoid split, which is named as ``HomiCerco''.
We know the most recent common ancestor of humans and chimps may live about $5-7$ million years ago, a normal distribution centered at 6 million years with a standard deviation of 0.5 million years is assigned to the ``Human-Chimp'' prior.
Another normal distribution of $24 \pm 0.5$ (million years) is likewise defined for ``HomiCerco''.

\begin{figure}[ht]
\subfigure[Estimates]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/treeRootHeight.pdf}}
\subfigure[Density]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/multiKDE.pdf}}\\
\subfigure[Joint-Marginal]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/joint-marginal-th-ingroup.pdf}}
\subfigure[Trace]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/trace.pdf}}
\caption{Images exported from four analysis tabs in Tracer for calibrating the tree to estimate TMRCA, where the black is ``treeModel.rootHeight'' parameter, blue is TMRCA parameter of ``Human-Chimp'', red is ``ingroup'', and yellow is ``HomiCerco''.}
\label{fig:4tabs}
\end{figure}

After loading the log, selecting the ``treeModel.rootHeight'' parameter gives the marginal posterior distribution of the age of the root of entire tree. The blue bars in Figure~\ref{fig:4tabs}.a present the most compact interval on the root height of the tree that contains 95\% of the posterior probability, referring to the period between about 39.01 and 63.67 million years old with the mean of 50.48 million years.
%TODO rate is faster than Yang's book r_ape = 5.4 * 10^-10 per year, r_monkey = 8.0, r_root = 6.6
Correspondingly selecting ``meanRate'' provides the rate of evolution averaged over the whole tree, which has the mean of 0.0129 substitutions per site per million years with 95\% HPD Interval between 0.0101 and 0.0159.

Figure~\ref{fig:4tabs}.b shows the kernel density estimations of these four parameters. The two TMRCA parameters, ``Human-Chimp'' in blue and ``HomiCerco'' in yellow, that we used to calibrate the tree explicitly have posterior distributions very similar to the prior distributions that we specified.

The marginal posterior densities of ``treeModel.rootHeight'' and ``ingroup'' are nearly overlapped each other. When we switch to their joint-marginal distribution plot in Figure~\ref{fig:4tabs}.c, the samples are almost lined up together, which indicates they are associated.
But this isn't an issue here, because the ``ingroup'' only excludes one outgroup species from all taxa.
%TODO how to explain this problem further?

Figure~\ref{fig:4tabs}.d illustrates the traces of these four parameters against their logged states from MCMC chain. ``Human-Chimp'' and ``HomiCerco'' parameters are immediately reaching equilibrium from the starting point, but ``treeModel.rootHeight'' and ``ingroup'' clearly show a burn-in period. %TODO ?

\subsection*{Non-parametric coalescent estimator}

In the second example, there are two analyses conducted using a H3N2 data set that is the subset of a comprehensive influenza A virus data set spanning several epidemic seasons in the New York state. The data contains a  ?bp alignment of 165 Hemagglutinin genes collected between about 1999 and 2004 \citep{rambaut2008genomic}, explicating H3N2 spreading during three epidemic seasons.

To be able to reconstruct Bayesian coalescent skygrid \citep{minin2008smooth} plot, we exploit Tracer to exact past population dynamics from both MCMC log and tree log produced by a BEAST \citep{drummond2012bayesian} run.
The HKY + $\Gamma_4$ model, strict molecular clock, and Bayesian skygrid tree prior was respectively setup in the run, where a grid of 50 intervals over 5 years estimating 10 population sizes by year was constructed in the skygrid prior.

%What type of dynamics does the H3N2 skyride plot suggest?
According to Tracer's skygrid plot in Figure~\ref{fig:flu}.a, the changing patterns of genetic diversity in viral isolates from New York state clearly reveal the seasonal dynamics of influenza A in individual temperate populations. The absolute amount of genetic diversity, even at seasonal peaks, is small compared to other rapidly evolving viruses that infect far fewer people, suggesting that strong natural selection, in addition to periodic bottlenecks, reduces the level of diversity that co-circulates at any time. Simulations demonstrate that this reconstruction of genetic diversity is robust to the sampling protocol. %TODO copied from (Rambaut et al., 2008)

\begin{figure}[ht]
%\includegraphics[width=.5\textwidth]{./figures/fluBSP.pdf}
%\subfigure[Bayesian coalescent skygrid]{\includegraphics[width = .23\textwidth]{./figures/fig21.png}}
\centerline{
\includegraphics[width=0.4\textwidth]{./figures/fig21.png}
}
% MAS: Please use vector-graphic format, like PDF
%\subfigure[Lineages through time]{\includegraphics[width = .23\textwidth]{./figures/fluLTT.pdf}}
\caption{(a) Bayesian coalescent skygrid plot and (b) lineages through time for the Influenza A/H3N2 virus data.}
\label{fig:flu}
\end{figure}

Lineages Through Time in Figure~\ref{fig:flu}.b, similar curve

diversification has been gradually developed before 2003's break-out, but it also looks stopped after 2004.

\section*{Availability and Future Directions}

We make the Tracer package available in both executable and source code forms.  Tracer requires Java version 1.6 or greater and executables for Windows, Mac OS and Linux platforms are located at \url{http://beast.bio.ed.ac.uk/Tracer} % \url{https://github.com/beast-dev/tracer} ?
which serves as the main page for the package. This page also links to a sizeable list of self-contained, step-by-step tutorials covering basic to advance usage of Tracer to summarise posterior distributions of a large set of phylogenetic models simulated using BEAST.  For example, popular tutorials describe how to use Tracer to generate marginal parameter summaries and infer population dynamics trajectories over time.

Github houses the Tracer's source code within \url{https://github.com/beast-dev/tracer} ,and links to a GoogleGroup discussion group related to Tracer,
which is the ``beast-users" group (\url{http://groups.google.com/group/beast-users}) with over 1,500 members.
%At the time of writing, forty-seven developers belong to the ``beast-dev" group that facilitates BEAST development across three continents.

Future development directions for Tracer focus on \ldots

@@

\section*{Acknowledgments}

This work was supported in part by
the Wellcome Trust through project 206298/Z/17/Z (Artic Network),
the European Union Seventh Framework Programme under grant agreement no. 725422-RESERVOIRDOCS,
the National Science Foundation through grant DMS 1264153,
the National Institutes of Health under grant R01 AI107034,
and
the Marsden Trust.

>>>>

\author[Rambaut \textit{et~al.}]{ Andrew Rambaut\,$^{1}$, Guy Baele$^{2}$,  Dong Xie\,$^{3,4}$, Marc A.~Suchard\,$^{5,6,7}$, Alexei J.~Drummond\,$^{3,4}$}

\address{
$^{1}$Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK\\
$^{2}$Department of Microbiology and Immunology, Rega Institute, KU Leuven, Leuven, Belgium\\
$^{3}$Department of Computer Science, University of Auckland, Auckland, NZ\\
$^{4}$Centre for Computational Evolution, University of Auckland, Auckland, NZ\\
$^{5}$Departments of Biomathematics and Human Genetics, David Geffen School of Medicine, University of California, Los Angeles, USA \\
$^{6}$Department of Biostatistics, UCLA Fielding School of Public Health, University of California, Los Angeles, USA \\
$^{7}$Department of Human Genetics, David Geffen School of Medicine at UCLA, University of California, Los Angeles, USA\\
}

@@

\section{Motivation:}
Bayesian inference of phylogeny using Markov chain Monte Carlo (MCMC) plays a central role in understanding evolutionary history and processes from molecular sequence data.
Visualising and analysing the samples from the posterior distribution is a key step in any non-trivial Bayesian inference, and it is especially important for methods that employ MCMC.
\section{Results:}
Here we present a software package \emph{Tracer} (version 1.7) for visualising and analysing the trace files generated through Bayesian inference.
The \emph{Tracer} software also provides kernel density estimation, multivariate visualisation, demographic trajectory reconstruction, conditional posterior distribution summary and more.
\section{Availability:}
Tracer is open-source under the GNU lesser general public license and available at \url{http://http://beast.community/tracer} and \url{https://github.com/beast-dev/tracer}.

@@

Bayesian inference of phylogeny using Markov chain Monte Carlo (MCMC) \citep{rannala1996probability, mau1999bayesian} flourishes as a popular approach to uncover the evolutionary relationships among operational taxonomic units (OTUs), such as genes, genomes, individuals or species.
MCMC approaches generate samples of model parameter values - including the phylogenetic tree - drawn from their posterior distribution given molecular sequence data and a selection of evolutionary models.
Visualising, tabulating and marginalising these samples is critical for approximating the posterior quantities of interest that one reports as the outcome of a Bayesian phylogenetic analysis.
To facilitate this task, we have developed the Tracer (version 1.7) software package, which allows users to process MCMC log files containing parameter samples and to interactively explore the high-dimensional posterior distribution.
Tracer works automatically with sample output from BEAST 1 \citep{drummond2012bayesian}, BEAST 2 \citep{bouckaert2014beast2}, MrBayes \citep{ronquist2012mrbayes}, RevBayes \citep{hohna2016revbayes}, LAMARC \citep{kuhner2006lamarc}, Migrate \citep{beerli2006comparison} and possibly other MCMC programs from other domains.

@@

Tracer is able to analyse a single MCMC log and can also combine posterior samples from multiple trace files stemming from an independent evaluation of the same dataset with an identical model combination.
For each trace file loaded, Tracer analyses the posterior samples from all the available parameters, regardless of whether these are discrete or continuous, and present statistical summaries as well as visualisation options for all of these parameters.
Importantly, the effective sample size (ESS) - an important measurement designed for a sampled parameter having continuous values - is immediately apparent in the default view of Tracer and allows users to assess the number of effectively independent draws from the posterior distribution for that parameter (see Figure \ref{fig:overview}:a).
Colour coding is provided for the ESS values to assist the user in determining potential mixing problems in the Bayesian inference performed, with cut-off values at 100 and 200.

Tracer is able to handle three trace types, also known as variable types \citep{mendenhall2012introduction}: real, integer, and categorical.
The real trace type is used to handle quantitative continuous variables, which commonly appear in phylodynamic analyses, whereas the integer trace type allows to deal with quantitative discrete variables and the categorical trace is meant for qualitative categorical variables.
The three trace types are automatically assigned when the trace file is imported but can also manually be set.

Multiple parameter traces can be selected from the traces panel, which will result in a side-by-side comparison or overlay of the selected parameters' visualisations for the different traces (see Figure \ref{fig:overview}:b-e). 
Multiple trace files can be selected in a similar fashion to compare posterior samples between different replicates of an analysis. 
If multiple trace files have the same file name,  then a ``Combined'' trace will automatically appear, which can be selected as well as the individual trace files.
Tracer consists of four panels that allow to present both individual and joint visualisation options for either a single or multiple selected trace files:

\begin{itemize}

\item Estimates: this panel reports common summary statistics such as the mean, standard deviation, highest posterior density interval, ESS, and other statistics concerning the selected parameter(s). 
A histogram presenting the frequency distribution over values will be shown for a single selected parameter of any trace type (see Figure \ref{fig:overview}:a), whereas side-by-side boxplots will be shown when multiple continuous parameters are selected (see Figure \ref{fig:overview}:b).

\item Marginal Density: this panel shows a density plot for the posterior sample of the selected parameter(s). Kernel density estimation (see Figure \ref{fig:overview}:c), histograms and violin plots (see Figure \ref{fig:overview}:d) are available for continuous traces, whereas frequency plots are employed for integer traces.

\item Joint-Marginal: visualisation in this panel only appears when two or more parameters are selected. The actual visualisation depends on the trace type of the selected parameters. We show several examples in the next section of the paper.

\item Trace: this panel shows a line plot connecting the posterior samples of one or more selected parameters against state or generation number (see Figure \ref{fig:overview}:e). Such a trace plot is typically used to assess mixing, select a suitable burn-in and identify trends that suggest convergence issues.

@@

Tracer offers an elegant solution of visualising conditional posterior distributions. 
A typical use case involves support for Bayesian stochastic search variable selection (BSSVS), a form of model averaging, in which parameters appear in the likelihood function only when a specific model is indicated by an indicator function.
When BSSVS is used, the posterior distribution of the parameter should not include those states that were sampled from the prior when its model was not part of the likelihood.
Discrete phylogeographic analyses frequently employ such an approach due to the potentially large amount of transition rates that need to be estimated \citep{Lemey2009}, but this is also relevant when employing model averaging approaches such as for relaxed molecular clocks for example \citep{Li2012}.

Additionally, Tracer provides demographic reconstruction resulting in a graphical plot, often applied to reconstruct epidemic dynamics.
Depending on the actual demographic model used in the analysis, an additional file containing samples from the posterior tree distribution will be required.
The currently available models are constant size, exponential and logistic growth \citep{drummond2002estimating}, Bayesian coalescent skyline \citep{drummond2005bayesian}, Gaussian Markov Random Field (GMRF) Bayesian skyride \citep{minin2008smooth}, and Bayesian coalescent skygrid \citep{gill2012improving}.
%Not sure if the statement below is correct ...
All of these demographic models are available in BEAST 1 \citep{drummond2012bayesian} and BEAST 2 \citep{bouckaert2014beast2}.
%Besides these, ``Lineages Through Time'' (?) is also available to plot the quantiles of the number of lineages against time, which describes the change of diversification of all organisms over time. %citation?

\begin{figure}[ht]
\subfigure[Tracer overview]{\includegraphics[width=0.47\textwidth]{./figures/rabv-overview.pdf}}
\subfigure[Boxplot]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-boxplots.pdf}}
\subfigure[Density]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-density.pdf}}\\
\subfigure[Violin]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-violin.pdf}}
\subfigure[Trace]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-traces.pdf}}
\caption{Overview of Tracer functionality and individual parameter visualisations: (a) Main Tracer panel upon loading a single trace file; (b) boxplot representation of two continuous parameters; (c) kernel density estimates; (d) violin plots; (e) the actual traces connecting the parameter values visited by the Markov chain.}
\label{fig:overview}
\end{figure}

@@

\subsection*{Cross-species dynamics of rabies virus in North American bats}

We provide on example on reconstructing the spatial dispersal and cross-species dynamics of rabies virus (RABV) in North American bat populations based on a set of 372 nucleoprotein gene sequences (nucleotide positions: 594-1353). 
The data set comprises a total of 17 bat species sampled between 1997 and 2006 across 14 states in the United States \citep{Streicker}, including two additional species that had been excluded from the original analysis owing to a limited amount of available sequences \citep{Faria2013}.
%We also include a viral sequence with an unknown sampling date (accession no. TX5275, sampled in Texas from Lasiurus borealis), which will be adequately accommodated in our inference.
We estimate the ancestral locations of RABV using a Bayesian discrete phylogeographic approach and, at the same time, infer the history of host jumping using the same model approach \citep{Lemey2009}. 
We employ the Bayesian skygrid coalescent model \citep{gill2012improving} to estimate the effective population sizes over time.

\begin{figure}[ht]
\subfigure[TangHuLu chart]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-integerjointmarginal.pdf}}
\subfigure[Marginal density]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-integermarginaldensity.pdf}} \\
\subfigure[Scatter plot]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-correlation.pdf}}
\subfigure[Correlation plot]{\includegraphics[width = .23\textwidth, height = 3cm]{./figures/rabv-joint-real.pdf}} \\
\subfigure[Box-and-whisker plot]{\includegraphics[width = .465\textwidth, height = 3cm]{./figures/rabv-catreal.pdf}}
\caption{Joint marginal and marginal density plots of: (a) two integer variables through a TangHuLu chart; (b) frequency plots the marginal density of multiple integer or categorical variables; (c) classic scatter or correlation plot for two continuous variables; (d) large correlation matrices for multiple ($> 2$) continuous variables; (e) box-and-whisker plots for the joint marginal distribution of a continuous and an integer or categorical variable.}
\label{fig:4tabs}
\end{figure}

Performing BSSVS in a discrete phylogeographic approach provides us with parameters of the integer and categorical trace type.
While a single parameter of such types can easily be visualised using a frequency plot, we have developed a TangHuLu chart to visualise the joint probability between two integer or categorical traces using coloured bubbles (see Figure \ref{fig:4tabs}:a).
The size of circle drawn is proportional to the joint probability, %the blue coloured circle is in the credible set of given a probability threshold default to 0.95, the red is in the set not in the credible set. 
with the tile background being coloured if there is in fact a circle, to enhance visibility. 
%The coloured background can also help to show the area of the credible set and non-credible set.
Multiple integer parameters can also be visualised using marginal density plots (see Figure \ref{fig:4tabs}:b), with a different colour scale being assigned to each parameter.
Both figures show a clear overlap in the number of non-zero transition rates in the discrete trait models for host and location, as identified using a BSSVS procedure.
The visualisation options for integer or categorical parameters are complemented with popular plots for continuous variables, such as a scatter or correlation plot (see Figure \ref{fig:4tabs}:c) for two continuous variables, which indicates a strong positive correlation between the clock rates of both discrete trait models. 
We also provide an extension of such a scatter plot for multiple (i.e. more than two) continuous variables (see Figure \ref{fig:4tabs}:d; \citet{Murdoch}), which makes use of coloured ellipses to indicate the strength of correlation between each pair of continuous variables.
Colour gradients indicate the strength and direction of the correlation, with a colour gradient from white to dark blue indicating increasingly strong correlation and a colour gradient from white to dark red indicating increasingly strong anti-correlation.
Additionally, the strength of correlation is reflected in the shape of the ellipse drawn, with no correlation being drawn as a circle and perfect (anti-)correlation being drawn as a line.
Finally, in order to display the joint marginal distribution between a continuous variable and an integer or categorical variable, we make use of box-and-whisker plots.
Using such a plot (Figure \ref{fig:4tabs}:e), we can infer that the range of the non-zero transition rates can depend on the ancestral location state inferred at the root of the underlying phylogeny.

The Bayesian skygrid coalescent model \citep{gill2012improving} allows us to use Tracer to reconstruct the demographic history of RABV in North America by drawing the effective population sizes over time (see Figure \ref{fig:rabv}).
Given the properties of this coalescent prior, which requires to specify a grid of intervals over the timeframe of the epidemic, Tracer is able to infer this history without requiring a posterior sample of trees.
According to Tracer's skygrid plot, it is clear that rabies virus has successfully established itself in North American bat species, with its numbers rising steadily throughout recent centuries.
Following a decline in numbers at the end of last century, we observe a recent sharp increase in population size of RABV in North American bats.

\begin{figure}[ht]
\centerline{
\includegraphics[width=0.46\textwidth]{./figures/rabv-skygrid.pdf}
}
\caption{Estimating the effective population sizes over time using a Bayesian skygrid demographic reconstruction for rabies virus in North America.}
\label{fig:rabv}
\end{figure}

\section*{Availability}

Tracer 1.7 is open-source under the GNU lesser general public license and available in both executable and source code forms at \url{http://beast.community/tracer}.
It requires Java version 1.6 or greater.
Github houses the Tracer's source code within \url{https://github.com/beast-dev/tracer}, and links to a GoogleGroup discussion group related to Tracer, i.e. the ``beast-users" group (\url{http://groups.google.com/group/beast-users}).
This page also links to a sizeable list of self-contained, step-by-step tutorials covering basic to advance usage of Tracer to summarise posterior distributions of a large set of phylogenetic models simulated using BEAST.  
For example, popular tutorials describe how to use Tracer to generate marginal parameter summaries and infer population dynamics trajectories over time.

@@

\section*{Acknowledgements}

This work was supported in part by the Wellcome Trust through project 206298/Z/17/Z (Artic Network), the European Union Seventh Framework Programme under grant agreement no. 725422-RESERVOIRDOCS, the National Science Foundation through grant DMS 1264153, the National Institutes of Health under grant R01 AI107034, and the Marsden Trust.
GB acknowledges support from the Interne Fondsen KU Leuven / Internal Funds KU Leuven.
